\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-11-10 parts-infer-ig.tex
ensure some notations are introduced

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
\modLP{The inference relies on finding configurations $\sigma$ of groups $g$ of regulators of $a$
such that the increase of one of the regulator $b\in g$ from $b_t$ to $b_{t+1}$
changes the direction of the evolution of $a$.
If $b\neq a$ (\pref{eq:edges-inference}), this is achieved by finding a configuration $\sigma$ of $g$ and $a$ such
that there exists a bounce $a_j$ from $a_i$ in $\sigma\{b_t\}$ which is different
from a bounce $a_k$ from $a_i$ in $\sigma\{b_{t+1}\}$.
If so, $b$ regulates $a$ at threshold $t+1$ with the sign of $k-j$.
The case where  $b = a$ (\pref{eq:edges-inference-auto}) is similar: we look for
$a_t$ such that there exists a bounce in opposite direction in $\sigma\{a_t\}$
than in $\sigma\{a_{t+1}\}$.
If there is no bounce in one of these two configurations, we also need to ensure
that the configuration corresponds to a local fixed point between all the
regulators of $a$ (instead of in scope of $g$).
The set of configuration where no bounce occur on $a_i$ is given given by
$\fp(a_i)$ (\pref{eq:fp}).
In \pref{eq:edges-inference} and \pref{eq:edges-inference-auto} we use
$\irB_a^g(\sigma)$ as a shortcut for $\focals$, which returns the set of
possible bounces of $a$ in the configuration $\sigma$.}
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp.
$\hat{E}_-$) by
\modLP{
\begin{align}
\begin{split}
\forall a\in\Gamma, \forall b\in\PHpredecgene{a}, b\neq a, & \forall s \in \{ +, - \}, \forall t < l_b\\
  b \xrightarrow{t+1} a \in \hat{E}_s \EQDEF & \exists g \in X(a), 
  	b \in g, \exists \sigma \in L(g\cup\{a\}), \\
    \qquad & \exists a_j \in \irB_a^g(\sigma\{b_t\}), \exists a_k \in \irB_a^g(\sigma\{b_{t+1}\}), \\
	\qquad & k > j \wedge s=+  \vee k < j \wedge s=-
\end{split}
\label{eq:edges-inference}
\\
\begin{split}
 \forall a\in\Gamma, & \forall s \in \{ +, - \}, \forall t < l_a\\
  a \xrightarrow{t+1} a \in \hat{E}_s \EQDEF & \exists g \in X(a), 
    \exists \sigma \in L(g\cup\{a\}), \\
  \qquad & \exists a_j \in \irB_a^{g \modMF{\cup \{a\}}}(\sigma\{a_t\}),
    \exists a_k \in \irB_a^{g \modMF{\cup \{a\}}}(\sigma\{a_{t+1}\}), \\
  \qquad & 
    (j=t \Rightarrow \exists \sigma'\in\fp(a_t): \sigma\{a_t\}\subseteq \sigma') \\
  \qquad & \wedge 
    (k=t+1 \Rightarrow \exists \sigma'\in\fp(a_{t+1}): \sigma\{a_{t+1}\}\subseteq \sigma') \\
  \qquad & \wedge (k\geq t+1 \wedge j \leq t \wedge s=+ \\
    \qquad & \vee k<t+1 \wedge j>t \wedge s=-)
\end{split}
\label{eq:edges-inference-auto}
\end{align}
\modMF{where $\irB_a^g(\sigma) \subset \PHl_a$ is defined as follows:}
\begin{align}
\bar\irB_a^g(\sigma) &\DEF \{ a_j \mid \PHfrappe{b_k}{a_i}{a_j}\in\PHa, b\in g \wedge b_k\in\ctx_a^g(\sigma)
\wedge a_i\in\sigma \}
\\
\irB_a^g(\sigma) &\DEF
\begin{cases}
\bar\irB_a^g(\sigma) & \text{if }\bar\irB_a^g(\sigma)\neq\emptyset \\
\{\PHget{\sigma}{a}\} & \text{otherwise}
\label{eq:irB}
\end{cases}
\end{align}
and $\fp(a_i) \subseteq  L(\PHpredecgene{a}\cup\{a\})$ is defined as follows:
\begin{equation}
\fp(a_i) \DEF \{\sigma\in L(\PHpredecgene{a}\cup\{a\}) \mid a_i\in\sigma \wedge
					\focals(a, \{a_i\}, \ctx_a^{\PHpredecgene{a}}(\sigma)) =
					\{a_i\} \}
\label{eq:fp}
\end{equation}
}
\end{proposition}

>>>>

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
\modLP{The inference relies on finding configurations $\sigma$ of groups $g$ of regulators of $a$
such that the increase of one of the regulator $b\in g$ from $b_t$ to $b_{t+1}$
changes the direction of the evolution of $a$.
If $b\neq a$ (\pref{eq:edges-inference}), this is achieved by finding a configuration $\sigma$ of $g$ and $a$ such
that there exists a bounce $a_j$ from $a_i$ in $\sigma\{b_t\}$ (\ie the state $\sigma$ where the
process of sort $b$ has been replaced with $b_t$) which is different
from a bounce $a_k$ from $a_i$ in $\sigma\{b_{t+1}\}$.
If so, $b$ regulates $a$ at threshold $t+1$ with the sign of $k-j$.
The case where  $b = a$ (\pref{eq:edges-inference-auto}) is similar: we look for
$a_t$ such that there exists a bounce in opposite direction in $\sigma\{a_t\}$
than in $\sigma\{a_{t+1}\}$.
If there is no bounce in one of these two configurations, we also need to ensure
that the configuration corresponds to a local fixed point between all the
regulators of $a$ (instead of in scope of $g$).
The set of configuration where no bounce occur on $a_i$ is given given by
$\fp(a_i)$ (\pref{eq:fp}).
In \pref{eq:edges-inference} and \pref{eq:edges-inference-auto} we use
$\irB_a^g(\sigma)$ as a shortcut for $\focals$, which returns the set of
possible bounces of $a$ in the configuration $\sigma$.}
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp.
$\hat{E}_-$) by
\modLP{
\begin{align}
\begin{split}
\forall a\in\Gamma, \forall b\in\PHpredecgene{a}, b\neq a, & \forall s \in \{ +, - \}, \forall t < l_b\\
  b \xrightarrow{t+1} a \in \hat{E}_s \EQDEF & \exists g \in X(a), 
  	b \in g, \exists \sigma \in L(g\cup\{a\}), \\
    \qquad & \exists a_j \in \irB_a^g(\sigma\{b_t\}), \exists a_k \in \irB_a^g(\sigma\{b_{t+1}\}), \\
	\qquad & k > j \wedge s=+  \vee k < j \wedge s=-
\end{split}
\label{eq:edges-inference}
\\
\begin{split}
 \forall a\in\Gamma, & \forall s \in \{ +, - \}, \forall t < l_a\\
  a \xrightarrow{t+1} a \in \hat{E}_s \EQDEF & \exists g \in X(a), 
    \exists \sigma \in L(g\cup\{a\}), \\
  \qquad & \exists a_j \in \irB_a^{g \modMF{\cup \{a\}}}(\sigma\{a_t\}),
    \exists a_k \in \irB_a^{g \modMF{\cup \{a\}}}(\sigma\{a_{t+1}\}), \\
  \qquad & 
    (j=t \Rightarrow \exists \sigma'\in\fp(a_t): \sigma\{a_t\}\subseteq \sigma') \\
  \qquad & \wedge 
    (k=t+1 \Rightarrow \exists \sigma'\in\fp(a_{t+1}): \sigma\{a_{t+1}\}\subseteq \sigma') \\
  \qquad & \wedge (k\geq t+1 \wedge j \leq t \wedge s=+ \\
    \qquad & \vee k<t+1 \wedge j>t \wedge s=-)
\end{split}
\label{eq:edges-inference-auto}
\end{align}
\modMF{where $\irB_a^g(\sigma) \subset \PHl_a$ is defined as follows:}
\begin{align}
\bar\irB_a^g(\sigma) &\DEF \{ a_j \mid \PHfrappe{b_k}{a_i}{a_j}\in\PHa, b\in g \wedge b_k\in\ctx_a^g(\sigma)
\wedge a_i\in\sigma \}
\\
\irB_a^g(\sigma) &\DEF
\begin{cases}
\bar\irB_a^g(\sigma) & \text{if }\bar\irB_a^g(\sigma)\neq\emptyset \\
\{\PHget{\sigma}{a}\} & \text{otherwise}
\label{eq:irB}
\end{cases}
\end{align}
and $\fp(a_i) \subseteq  L(\PHpredecgene{a}\cup\{a\})$ is defined as follows:
\begin{equation}
\fp(a_i) \DEF \{\sigma\in L(\PHpredecgene{a}\cup\{a\}) \mid a_i\in\sigma \wedge
					\focals(a, \{a_i\}, \ctx_a^{\PHpredecgene{a}}(\sigma)) =
					\{a_i\} \}
\label{eq:fp}
\end{equation}
}
\end{proposition}

### 2014-11-10 parts-tr2global.tex
minor corrections
ensure some notations are introduced

\modLP{%
As described in the introduction of this section, the identification of Thomas
models from PH relies on the inference of focal processes in the
different configurations of the regulators of each component.
In order to reduce the scope of the configurations that need to be enumerated, 
we introduce the notion of \emph{group of regulators} of sort $a$ as a set of
components that have a joint influence on $a$.
More precisely, $b$ and $c$ are in the same group of regulators of $a$ if there exists
an intermediate cooperative sort $\upsilon$ that hits $a$ such that $b$ and $c$
are regulators of $\upsilon$.
Those groups form a partition of $\PHpredecgene{a}$.
We denote $X(a)$ the finest partition in groups of
regulators (\pref{def:influence-groups}), i.e. where each group is minimal.
\begin{definition}[Partition $X(a)$ of $\PHpredecgene{a}$]
\label{def:influence-groups}
Given $a\in\PHs$,
$X(a)$ is the finest partition of $\PHpredecgene{a}$ such that
for any $b,c\in\PHpredecgene{a}$, $b\neq c$, if there exists a cooperative sort
$\upsilon\in\PHdirectpredec{a}\cap\CSorts$ such that 
$\{b,c\}\subseteq\PHpredecgene{\upsilon}$ then
there exists $g\in X(a)$ such that $\{b,c\}\subseteq g$.
If $\PHpredecgene{a} = \emptyset$, $X(a)\DEF \{\emptyset\}$.
\end{definition}
}

\modLP{%
If the cooperative sorts are all deterministic (\pref{def:deterministic-sort}),
it is sufficient to specify the configuration $\sigma$ of a group $g$ of regulators of a sort $a$
to obtain the configuration of all the sorts hitting $a$ directly.
Indeed, because of the absence of cycles between cooperative sorts
(\pref{pro:wf-ph}), one can recursively
evaluate the focal process of each cooperative sort that hit $a$ with respect to
the configuration $\sigma$ of the regulators $g$.
Such an extended configuration is denoted by $\ctx_a^g(\sigma)$, formalized in
\pref{def:ctx-sigma}
\begin{definition}[Configuration extension $\ctx_a^g(\sigma)$]
\label{def:ctx-sigma}
Given $a\in\PHs$, a sub-set of regulators $g\subseteq \PHpredecgene{a}$ and a configuration 
$\sigma\in L(g)$, $\ctx_a^g(\sigma)$ is the configuration $\sigma$ with the corresponding
processes of the cooperative sorts in $\PHdirectpredec{a}$:
\begin{align}
\ctx_a^g(\sigma) & \DEF \sigma \uplus \state{\upsilon(\sigma) \mid\upsilon \in\PHdirectpredec{a} \cap \CSorts} 
\label{eq:ctx-sigma}
\\
\upsilon(\sigma') & \DEF
\begin{cases}
\focals(\upsilon,L_\upsilon,\sigma') & \text{if
$\PHdirectpredec{\upsilon}\subseteq\f{dom}(\sigma')$}
\\
\upsilon(\sigma' \uplus \state{\upsilon'(\sigma') \mid 
  \upsilon'\in\PHdirectpredec{\upsilon} \cap \CSorts })
  & \text{otherwise,}
\end{cases}
\label{eq:cooperative-eval}
\end{align}
where $\f{dom}(\sigma')$ is the set of sorts defined in $\sigma'$.
\end{definition}
}

>>>>

\modLP{%
As described in the introduction of this section, the identification of Thomas
models from PH relies on the inference of focal processes in the
different configurations of the regulators of each component.
In order to reduce the scope of the configurations that need to be enumerated, 
we introduce the notion of \emph{group of regulators} of sort $a$ as a set of
components that have a joint influence on $a$.
More precisely, $b$ and $c$ are in the same group of regulators of $a$ if there exists
an intermediate cooperative sort $\upsilon$ that hits $a$ such that $b$ and $c$
are regulators of $\upsilon$.
Those groups form a partition of $\PHpredecgene{a}$.
We denote $X(a)$ the finest partition in groups of
regulators (\pref{def:influence-groups}), \ie where each group is minimal.
\begin{definition}[Partition $X(a)$ of $\PHpredecgene{a}$]
\label{def:influence-groups}
Given $a\in\PHs$,
$X(a)$ is the finest partition of $\PHpredecgene{a}$ such that
for any $b,c\in\PHpredecgene{a}$, $b\neq c$, if there exists a cooperative sort
$\upsilon\in\PHdirectpredec{a}\cap\CSorts$ such that 
$\{b,c\}\subseteq\PHpredecgene{\upsilon}$ then
there exists $g\in X(a)$ such that $\{b,c\}\subseteq g$.
If $\PHpredecgene{a} = \emptyset$, $X(a)\DEF \{\emptyset\}$.
\end{definition}
}

\modLP{%
If the cooperative sorts are all deterministic (\pref{def:deterministic-sort}),
it is sufficient to specify the configuration $\sigma$ of a group $g$ of regulators of a sort $a$
to obtain the configuration of all the sorts hitting $a$ directly.
Indeed, because of the absence of cycles between cooperative sorts
(\pref{pro:wf-ph}), one can recursively
evaluate the focal process of each cooperative sort that hit $a$ with respect to
the configuration $\sigma$ of the regulators $g$.
Such an extended configuration is denoted by $\ctx_a^g(\sigma)$, formalized in
\pref{def:ctx-sigma}
\begin{definition}[Configuration extension $\ctx_a^g(\sigma)$]
\label{def:ctx-sigma}
Given $a\in\PHs$, a sub-set of regulators $g\subseteq \PHpredecgene{a}$ and a configuration 
$\sigma\in L(g)$, $\ctx_a^g(\sigma)$ is the configuration $\sigma$ with the corresponding
processes of the cooperative sorts in $\PHdirectpredec{a}$:
\begin{align}
\ctx_a^g(\sigma) & \DEF \sigma \uplus \state{\upsilon(\sigma) \mid\upsilon \in\PHdirectpredec{a} \cap \CSorts} 
\label{eq:ctx-sigma}
\\
\upsilon(\sigma') & \DEF
\begin{cases}
\focals(\upsilon,L_\upsilon,\sigma') & \text{if
$\PHdirectpredec{\upsilon}\subseteq\f{dom}(\sigma')$}
\\
\upsilon(\sigma' \uplus \state{\upsilon'(\sigma') \mid 
  \upsilon'\in\PHdirectpredec{\upsilon} \cap \CSorts })
  & \text{otherwise,}
\end{cases}
\label{eq:cooperative-eval}
\end{align}
where $\f{dom}(\sigma')$ is the set of sorts defined in $\sigma'$
and $\uplus$ denotes the union of two (sub)states.
\end{definition}
}
